# Tibbles

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(magrittr)
```

## Introduction

Tibbles are an improvement over data frames. They have a similar API, that allows the easy use of list-columns.

```{r}
lst <- setNames(lapply(seq(1, 9, 3), function(i) i:(i + 2)), 
  c("A", "B", "C"))
tb1 <- tibble(x = 1:3, y = lst)
tb1
```

List-columns can be accessed with the usual data frame notation, although care should be taken when we want to access the values, rather than a sub-list.

```{r list_columns}
tb1[[2]]    # Returns the second columns, i.e. the list
tb1[1, 2]   # Returns a sub-list
tb1[[1, 2]] # Return the vector contained in the sub-list
```

List-columns are generally created by `do`. In this case we have created them manually.

### Name checking

By default tibbles do not check the names

```{r}
tb2 <- tibble("A re@lly *Bad* Name" = 1:3)
names(tb2)
```

### Lazy evaluation of columns
Columns in a tibble can be based on existing ones, thanks to lazy evaluation.

```{r}
tb3 <- tibble(
  x = seq(0, 2*pi, length.out = 10),
  y = sin(x)
)
plot(tb3)
```

### Row names
Row-names are deprecated, since all information should have its own column. Subsetting throws away attributes and this is often the cause of nasty bugs.

### Recycling
It only recycles vectors of lenght 1. Recycling can be powerful, but is a frequent cause of problems.

### Coercion

To coerce objects, in particular lists, into tibbles use `as_tibble`.
It doesn't change strings into factors and is much faster, especially when combining long lists into one object.

```{r}
ll <- list(a = 1:3, b = 4:6, c = LETTERS[1:3])
str(as.data.frame(ll))
str(as_data_frame(ll))
```

## Examples

### Adding new rows and columns

```{r add_row_column}
df <- tibble(x = 1:3, y = 3:1)
df %>% add_column(z = letters[1:3], w = LETTERS[3:1])
df %>% add_row(x = c(9, 10), y = c(10, 9))
```

### Equality

```{r all_equal}
df1 <- df
df2 <- df[c(3, 1, 2), ] # Scrambled version of df1
all_equal(df1, df2)
```

### Coercion

Coercion automatically adds suitable column names via `repair_names`.

```{r as_tibble}
m <- matrix(1:16, nrow = 4)
as_tibble(m)
```

### Finding columns

`has_name` is a convenient way to test for the presence of one or more column names.

```{r has_name}
df <- tibble(col1 = 1:3, col2 = 3:1, col3 = c('a', 'b', 'c'))
has_name(df, c('col1', 'col3'))
```

### Rownames to columns

`rowname_to_column` and `column_to_rowname` are two handy functions.

```{r rowname_to_and_from_column}
df %<>% column_to_rownames(var = "col3")
df
rownames_to_column(df, var = "row_names")
```

