# Window functions in `dplyr`

We use the `Lahman` package, containing the database of pitching, hitting and fielding statistics for Major League Baseball from 1871 to 2015. The database is composed of the following main tables:

- Master: player names, dates of birth, death, etc.
- Batting: batting statistics.
- Pitching: pitching statistics
- Fielding: fielding statistics.

Other tables are available, but we don't consider them here.

```{r}
library(tidyverse)
library(Lahman)
```

The `Batting` table contains `r nrow(Batting)` rows and `r ncol(Batting)` columns. Below we select the columns `playerID`, `yearID`, `teamID`, `G` (games, number of games in which a player played), `AB` (at bats), `R` (runs), `H` (hits).

```{r}
batting <- select(tbl_df(Batting), playerID, yearID, teamID, G, AB:H) 
batting <- arrange(batting, playerID, yearID, teamID)
players <- group_by(batting, playerID)
```

Note that we order by `playerID` and then by `yearID` so that, within a `playerID`, the `yearID` is already sorted. This ordering is important, since `lead` and `lag` know nothing about the columns on which they are applied, and they just apply positionally. In other words, a lag-1 operation will look at the previuos element in the same column, irrespective of whether it is sorted or not.
For example:

```{r}
tb <- tibble::tibble(x = c(1, 3, 4, 2, 6, 5))
filter(tb, x > lag(x))
```

The filtering leaves only the elements that are greater than the element that precedes them.

Going back to the `player` dataset, it has the following characteristics:

- `batting`: `r nrow(batting)` rows and `r ncol(batting)` columns.
- `players`: `r nrow(players)` rows and `r ncol(players)` columns.

## Types of Window Functions

### Windowed rank functions

There are six variations built upon base R's `rank`. The first three return integer ranks.

1. `row_number` does not break ties.
2. `min_rank` breaks ties creating gaps.
3. `dense_rank` breaks ties without creating gaps.

For example, using the same example as in the vignette.

```{r integer_ranking}
x <- c(1, 1, 2, 2, 2)
row_number(x)
min_rank(x)
dense_rank(x)
dense_rank(desc(x))
```

More in detail, let's see how `row_number` behaves in the following situations:

```{r row_number}
set.seed(123)
x1 <- c(2, 1, 1, 2, 2, 3, 1, 3, 3)
row_number(x1)
x2 <- sample(1:10)
x2
row_number(x2)
```

In the second case there are no ties, while in the first case there are several.

```{r ties_management}
row_number(x1)
min_rank(x1)
dense_rank(x1)
```

If we want to reverse the ranking, we just need to use `desc`, as shown in the last example.

Three more functions return a value between 0 and 1.

1. `percent_rank` rescales `min_rank` between 0 and 1.
2. `cume_dist` is the proportion of values less than or equal to the current rank.
3. `ntile` breaks the input vector into buckets.


```{r percent_ranking}
percent_rank(x)
percent_rank(desc(x))
cume_dist(x)
ntile(x, 3)
```

Note that `cume_dist` is the cumulative distribution function. It is very different from `percent_rank`. The `ntile` function returns a vector with the bin number each values falls in. 

### Example with the `players` dataset

The `players` dataset is grouped subset of the `Batting` dataset. The meaning of the non-obvious variables is:

- G: Games. Number of games played.
- AB: At bats.
- R: Runs.
- H: Hits. Times reached base.

If we want to know the three years in which each player played the largest number of games:

```{r three_years_most_games, eval=FALSE}
players %>% filter(min_rank(desc(G)) <= 3)
```



