# Standard Evaluation in `dplyr`

```{r}
library(tidyverse)
```

Using standard evaluation (SE for short) can be tricky. The most common case is when we have one or more column names stored in a variable, and we want to apply a function to it. The Non-standard evaluation vignette suggests three approaches:

1. Model formulas.
2. Quoting.
3. Strings.

## Selection

The following approaches all fail.

```{r, eval = FALSE}
my_col <- "mpg"
my_col_range <- "mpg:disp"
mtcars %>% select_(~my_col) # Fails
mtcars %>% select_(quote(my_col)) # Fails
```

These ones all succeed.

```{r, eval=FALSE}
mtcars %>% select_(my_col) # Succeeds
mtcars %>% select(as.name(my_col)) # Succeds
mtcars %>% select_(my_col_range) # Also succeeds, which is nice!
```

As a side note, if we want to select from a column up to the last one, the `last` function comes handy.

```{r to_the_last}
my_col <- "qsec"
my_col_range <- paste(my_col, last(names(mtcars)), sep = ":")
mtcars %>% select_(my_col_range)
```

Similarly if we want to start from the `first` column.

## Summarise

We usually face the following problems:

1. Summarising multiple variables with one function.
2. Summarising one variable with multiple functions.
3. Summarising multiple variables with multiple functions.

### Summarising multiple variables with one function

We store the names of the variables we want to summarise in a vector

```{r case_1}
my_cols <- c("mpg", "hp", "wt")
mtcars %>% select(one_of(my_cols)) %>% summarise_all(mean)
```

Note that we have called `select` and not `select_`, and we have used the `one_of` function to select the columns. If we try `select_` we obtain the wrong result.

```{r case_1_wrong}
mtcars %>% select_(my_cols) %>% summarise_all(mean)
```

### Summarising one variable with multiple functions

In the NSE case this is straightforward, we just need to specify each function individually.

```{r case2_nse, eval=FALSE}
mtcars %>% summarise(mean_mpg = mean(mpg), sd_mpg = sd(mpg))
```

In the SE case:

```{r case2_se, eval=FALSE}
mtcars %>% summarise_(~mean(mpg), ~sd(mpg)) # Succeeds
mtcars %>% summarise_("mean(mpg)", "sd(mpg)") # Succeeds
dots <- list("mean(mpg)", "sd(mpg)")
mtcars %>% summarise_(.dots = dots) # Also succeeds
```

In each case we must build the expressions "by hand". In other words, we cannot do something like:

```{r case2_nope, eval=FALSE}
my_col <- "mpg"
mtcars %>% summarise_(mean(my_col)) # Fails
```

### Summarising multiple variables with multiple functions

In the NSE case this works without problems if we use `funs` to create a list of functions and apply it with `summarise_all`.

```{r}
mtcars %>% select(mpg, cyl, disp) %>% summarise_all(funs(mean = mean, stdev = sd))
```

What *does not* work is applying a function that returns multiple values.

```{r case_3, eval = FALSE}
mtcars %>% select(mpg, cyl) %>% summarise_all(summary)
```

## Worked example: renaming the column of a tibble with the content of a variable

Let's suppose we have a tibble and we want to rename one of its columns, but the new name is stored in a variable. 

```{r worked_example}
dfr <- tibble(col_a = 1:4, col_b = 4:1)
new_col_name <- "col_x"
```

If we try to assign the column name directly using `rename_`, we will fail.

```{r, eval=FALSE}
dfr %>% rename_(new_col_name = col_x) # Fails
```
The `rename` and `rename_` functions expect named arguments. We can create one using `setNames`. The trick to make things work is to use the `.dots` argument as shown below.

```{r}
dfr %>% rename_(.dots = setNames("col_a", new_col_name))
```

